Loop-Aware Optimizations in PyPy's Tracing JIT

### 2011-06-21 talk/iwtc11/paper.tex
rename in benchmark implementations aswell
Fix my affiliation
add a small note about numpy and prolog

\authorinfo{H\aa kan Ardö}
           {Centre for Mathematical Sciences, Lund University}
           {hakan@debian.org}
\authorinfo{Carl Friedrich Bolz}
           {Heinrich-Heine-Universität Düsseldorf}
           {cfbolz@gmx.de}
\authorinfo{Maciej Fijałkowski}
           {Affiliation2}
           {fijall@gmail.com}

@@

XXX we either need to explain that we use C++ or consistently use C
\begin{figure*}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|r|r|r|}
\hline
 & CPython & Psyco & PyPy no LP & PyPy & GCC -O2 & GCC -O3\\
\hline
%NoBorder & 537.31 & 329.98 & 2.22 $\pm$ 0.03 & 2.17 $\pm$ 0.02 & - & -\\
%\hline
%NoBorder(iter) & 548.32 & 304.13 & 1.45 $\pm$ 0.03 & 1.47 $\pm$ 0.02 & - & -\\
%\hline
%NoBorder(range) & 534.64 & 317.34 & 1.34 $\pm$ 0.03 & 1.40 $\pm$ 0.04 & - & -\\
%\hline
conv3x3(NoBorderPadded(1000x1000)) & 543.73 & 333.20 & 2.09 $\pm$ 0.12 & 1.93 $\pm$ 0.05 & - & -\\
\hline
%NoBorderPadded(iter) & 546.70 & 309.32 & 1.21 $\pm$ 0.02 & 0.49 $\pm$ 0.02 & - & -\\
%\hline
%NoBorderPadded(range) & 550.92 & 318.33 & 1.12 $\pm$ 0.03 & 0.48 $\pm$ 0.01 & - & -\\
%\hline
conv3(array(1e5)) & 77.89 & 9.52 & 1.77 $\pm$ 0.06 & 0.68 $\pm$ 0.02 & 0.70 $\pm$ 0.05 & 0.59 $\pm$ 0.01\\
\hline
conv3(array(1e6)) & 77.15 & 9.58 & 1.69 $\pm$ 0.01 & 0.77 $\pm$ 0.01 & 0.84 $\pm$ 0.05 & 0.74 $\pm$ 0.01\\
\hline
conv3x3(Array2D(1000x1000)) & 23.72 & 12.77 & 0.07 $\pm$ 0.00 & 0.05 $\pm$ 0.03 & 0.24 $\pm$ 0.00 & 0.25 $\pm$ 0.01\\
\hline
conv3x3(Array2D(1000x3)) & 23.85 & 12.77 & 0.10 $\pm$ 0.00 & 0.07 $\pm$ 0.00 & 0.27 $\pm$ 0.01 & 0.27 $\pm$ 0.01\\
\hline
conv5(array(1e5)) & 122.54 & 16.67 & 1.86 $\pm$ 0.02 & 1.05 $\pm$ 0.03 & 1.03 $\pm$ 0.05 & 0.65 $\pm$ 0.01\\
\hline
conv5(array(1e6)) & 125.77 & 16.80 & 1.92 $\pm$ 0.03 & 1.09 $\pm$ 0.02 & 1.07 $\pm$ 0.05 & 0.80 $\pm$ 0.01\\
\hline
dilate3x3(Array2D(1000x1000)) & 23.29 & 12.99 & 0.41 $\pm$ 0.04 & 0.39 $\pm$ 0.01 & 0.26 $\pm$ 0.00 & 0.26 $\pm$ 0.01\\
\hline
sobel(NoBorderPadded(1000x1000)) & 461.14 & 258.02 & 1.01 $\pm$ 0.03 & 0.48 $\pm$ 0.03 & - & -\\
\hline
sobel(Array2D(1000x1000)) & - & - & - & - & 0.19 $\pm$ 0.01 & 0.20 $\pm$ 0.01\\
\hline
%sobel\_uint8(NoBorderPadded) & 476.72 & 275.50 & 1.05 $\pm$ 0.01 & 0.51 $\pm$ 0.00 & - & -\\
%\hline
sqrt(Fix16) & 744.35 & 421.65 & 3.93 $\pm$ 0.11 & 2.14 $\pm$ 0.02 & 0.97 $\pm$ 0.02 & 0.96 $\pm$ 0.01\\
\hline
sqrt(float) & 24.21 & 5.52 & 1.36 $\pm$ 0.03 & 1.00 $\pm$ 0.00 & 0.98 $\pm$ 0.00 & 0.98 $\pm$ 0.00\\
\hline
sqrt(int) & 20.84 & 1.78 & 2.26 $\pm$ 0.01 & 1.82 $\pm$ 0.01 & 0.81 $\pm$ 0.01 & 0.80 $\pm$ 0.01\\
\hline

@@

We can observe that PyPy (even without loop peeling) is orders of magnitude
faster than either CPython or Psyco. This is due to the JIT compilation
advantages and optimizations we discussed in XXX [ref to other paper]. Loop
peeling gives an additional XXX on average, which makes benchmark times
comparable with native-compiled C code. Missing performance we attribute to
the relative immaturity of PyPy's JIT assembler backend as well as missing
optimizations, like instruction scheduling. Results:

XXX add a small note somewhere that numpy and prolog are helped by this
optimization

>>>>

\authorinfo{H\aa kan Ardö}
           {Centre for Mathematical Sciences, Lund University}
           {hakan@debian.org}
\authorinfo{Carl Friedrich Bolz}
           {Heinrich-Heine-Universität Düsseldorf}
           {cfbolz@gmx.de}
\authorinfo{Maciej Fijałkowski}
           {Unaffiliated}
           {fijall@gmail.com}

@@

XXX we either need to explain that we use C++ or consistently use C
\begin{figure*}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|r|r|r|}
\hline
 & CPython & Psyco & PyPy no LP & PyPy & GCC -O2 & GCC -O3\\
\hline
%NoBorder & 537.31 & 329.98 & 2.22 $\pm$ 0.03 & 2.17 $\pm$ 0.02 & - & -\\
%\hline
%NoBorder(iter) & 548.32 & 304.13 & 1.45 $\pm$ 0.03 & 1.47 $\pm$ 0.02 & - & -\\
%\hline
%NoBorder(range) & 534.64 & 317.34 & 1.34 $\pm$ 0.03 & 1.40 $\pm$ 0.04 & - & -\\
%\hline
conv3x3(NoBorderPadded(1000x1000)) & 543.73 & 333.20 & 2.09 $\pm$ 0.12 & 1.93 $\pm$ 0.05 & - & -\\
\hline
%NoBorderPadded(iter) & 546.70 & 309.32 & 1.21 $\pm$ 0.02 & 0.49 $\pm$ 0.02 & - & -\\
%\hline
%NoBorderPadded(range) & 550.92 & 318.33 & 1.12 $\pm$ 0.03 & 0.48 $\pm$ 0.01 & - & -\\
%\hline
conv3(array(1e5)) & 77.89 & 9.52 & 1.77 $\pm$ 0.06 & 0.68 $\pm$ 0.02 & 0.70 $\pm$ 0.05 & 0.59 $\pm$ 0.01\\
\hline
conv3(array(1e6)) & 77.15 & 9.58 & 1.69 $\pm$ 0.01 & 0.77 $\pm$ 0.01 & 0.84 $\pm$ 0.05 & 0.74 $\pm$ 0.01\\
\hline
conv3x3(Array2D(1000x1000)) & 23.72 & 12.77 & 0.07 $\pm$ 0.00 & 0.05 $\pm$ 0.03 & 0.24 $\pm$ 0.00 & 0.25 $\pm$ 0.01\\
\hline
conv3x3(Array2D(1000000x3)) & 23.85 & 12.77 & 0.10 $\pm$ 0.00 & 0.07 $\pm$ 0.00 & 0.27 $\pm$ 0.01 & 0.27 $\pm$ 0.01\\
\hline
conv5(array(1e5)) & 122.54 & 16.67 & 1.86 $\pm$ 0.02 & 1.05 $\pm$ 0.03 & 1.03 $\pm$ 0.05 & 0.65 $\pm$ 0.01\\
\hline
conv5(array(1e6)) & 125.77 & 16.80 & 1.92 $\pm$ 0.03 & 1.09 $\pm$ 0.02 & 1.07 $\pm$ 0.05 & 0.80 $\pm$ 0.01\\
\hline
dilate3x3(Array2D(1000x1000)) & 23.29 & 12.99 & 0.41 $\pm$ 0.04 & 0.39 $\pm$ 0.01 & 0.26 $\pm$ 0.00 & 0.26 $\pm$ 0.01\\
\hline
sobel(NoBorderPadded(1000x1000)) & 461.14 & 258.02 & 1.01 $\pm$ 0.03 & 0.48 $\pm$ 0.03 & - & -\\
\hline
sobel(Array2D(1000x1000)) & - & - & - & - & 0.19 $\pm$ 0.01 & 0.20 $\pm$ 0.01\\
\hline
%sobel\_uint8(NoBorderPadded) & 476.72 & 275.50 & 1.05 $\pm$ 0.01 & 0.51 $\pm$ 0.00 & - & -\\
%\hline
sqrt(Fix16) & 744.35 & 421.65 & 3.93 $\pm$ 0.11 & 2.14 $\pm$ 0.02 & 0.97 $\pm$ 0.02 & 0.96 $\pm$ 0.01\\
\hline
sqrt(float) & 24.21 & 5.52 & 1.36 $\pm$ 0.03 & 1.00 $\pm$ 0.00 & 0.98 $\pm$ 0.00 & 0.98 $\pm$ 0.00\\
\hline
sqrt(int) & 20.84 & 1.78 & 2.26 $\pm$ 0.01 & 1.82 $\pm$ 0.01 & 0.81 $\pm$ 0.01 & 0.80 $\pm$ 0.01\\
\hline

@@

We can observe that PyPy (even without loop peeling) is orders of magnitude
faster than either CPython or Psyco. This is due to the JIT compilation
advantages and optimizations we discussed in XXX [ref to other paper]. Loop
peeling gives an additional XXX on average, which makes benchmark times
comparable with native-compiled C code. Missing performance we attribute to
the relative immaturity of PyPy's JIT assembler backend as well as missing
optimizations, like instruction scheduling.

Other interesting interpreters that are helped greatly by this
optimization are for
example our Prolog interpreter written in RPython, as well as numerical
kernel used for array manipulation. The exact extent is out of scope for
this paper.
